---
title: "Locational Marginal Price and Congestion Analysis in R"
authors: "Kate Esbenshade, Luke McFaul, Mack Smith"
format: revealjs  # Can also use ioslides, beamer, etc.
revealjs:
  theme: solarized
  slideNumber: true
  css: styles.css
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Global knitr chunk options knitr::opts_chunk$set(echo = FALSE, 
# Hide code by default warning = FALSE, 
# Suppress warnings message = FALSE 
# Suppress messages )
```

## Project Inspiration

1.  California has a unique electricity market structure that is
    different from other states in the US.
2.  There are many **integrated renewables** and **high fluctuation** in
    electricity demand, meaning that there are often times when there is
    not enough transmission capacity to move electricity from one
    location to another.
3.  This can cause **congestion in the transmission system**, which can
    lead to higher electricity prices in certain locations (different
    locational marginal prices, or LMPS).
4.  We wanted to create a project to **visualize** these differences in
    price and allow users to **understand and analyze** LMP differences
    and congestion.

## Project Goals

-   Create a package in R to help with analysis of electricity prices in
    California
-   Use the package to analyze the relationship between congestion and
    locational marginal prices (LMPs) in California wihin locations and
    between them
-   Create a shiny app to visualize the results

## Project Outputs

-   API wrapper for the gridstatus.io API to fetch and clean data of
    locational marginal prices for California between 2020-2023
-   A Shiny App that visualizes the transmission lines and locational
    data
-   Analysis functions for a single instance (15 minute interval) across
    all nodes, and for an entire node (or comparing nodes) across all
    instances of a year

## Example Plot from Package

This plot demonstrates the distribution of LMPs for an example node:

```{r visualize_node_example, echo=FALSE, fig.cap="LMP Distribution Example Plot"} 
# Load the CSV file 
# file_path <- system.file("extdata", "San_Francisco_node_2023.csv", package = "CAISOWrapR") 
# example_data <- read.csv(file_path)
```

## Generate and display the plot
